---
title: "NBA Data Analysis Project"
author: "Kate Li, Runzhe Li, Yifan Zhang, Linda Zhou"
date: "November 15, 2019"
output:
  html_document:
    highlight: tango
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
permalink: /description
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,warning=F, message=F)
```

## Motivation and Overview
The National Basketball Association (NBA) is a professional basketball league in North America that draws the attention of many basketball fans all over the world. During the regular season, each of the 30 teams (15 in the Eastern Conference and 15 in the Western Conference) plays 82 games. At the conclusion of the regular season, the top eight teams in each conference advance to the playoffs for a chance to compete for the championship title. Towards the end of the regular season, the race for the final few playoff spots are very intense, and usually it isn't until the final game of the regular season is over when we know which teams have made the cut. As NBA fans ourselves, we are eager to find out which teams will make the playoffs this year (2019-20 NBA season), but rather than wait until next April, as data analysts, we thought it would be interesting if we could make such predictions now. We hope our analysis will lead to meaningful insights that will be of interest to NBA fans all over the world. 

## Project Objectives
The goal of this project is to build a machine learning model that can predict the winning percentages of the NBA teams in the Western Conference, and thus the playoff teams. 

We would like to learn more about feature engineering and machine learning models. We will undergo a comprehensive background study of NBA to manually decide what features to use. Then feature learning techniques will be applied. We also plan to evaluate different models. A successful model will predict the 2019-2020 NBA playoffs in the Western Conference. Ideally the model could also be trained to predict the playoff teams in the Eastern Conference and finally round-by-round playoff predictions.

Some optional features we can consider include predictions of winning percentages for home games and for away games for each team.




## Related Work and Literature Review

### Playoff Prediction 
https://projects.fivethirtyeight.com/2020-nba-predictions/

### Pair-comparison data
Bradeley-Terry model

## Data 

### Data Source
The data will be collected from Basketball Reference website. The database contains player statistics since 1946, including points per game, assists per game, rebounds per game, etc,. The database also provides team statistics, including 3 point percentages, winning percentages, etc,. 

### Data Scraping 
We used `rvest` package to scrape the following data from Basketball Reference website:

* `team_stats`: The team statistics of first 20 games from season 2016-17 to 2019-20. We would use these data to define the team performance during specific season in the feature engineering section.

* `game_stats`: The game results (W-L record and scores) for each of the remaining 62 games from season 2016-17 to 2018-19. We would use these data in the mode building section. 

```{r,warning=F,message=F}
team_stats = readr::read_csv("./data/team_stats_all.csv")
game_stats = readr::read_csv("./data/game_stats_all.csv")
head(team_stats)
head(game_stats)

```

We did not include the code for web scraping in this rmarkdown file in order to save the runtime. You can check the `data_scraping.R` file in the source folder for more details. Here we just showed the first few rows of the data. 

### Data Wrangling
Data wrangling would be performed throughout the whole data analysis process covering data tidying, EDA, feature engineering, and model building. Here we demonstrated some necessary data wrangling procedures which were prepared for further analysis. 

```{r,warning=F,message=F}
library(rvest)
library(tidyverse)

west_ind = data.frame(
  Tm = c("CHI","LAL","POR","WAS","CHO","MIL","BRK","DET","HOU","NOP","IND","MEM","DEN","LAC","ATL","NYK","MIA","ORL","DAL","PHO",
         "SAC","UTA","MIN","SAS","CLE","TOR","BOS","PHI","GSW","OKC"),
  western = c(0,1,1,0,0,0,0,0,1,1,0,1,1,1,0,0,0,0,1,1,1,1,1,1,0,0,0,0,1,1),
  TeamName = c("Chicago Bulls","Los Angeles Lakers","Portland Trail Blazers", "Washington Wizards","Charlotte Hornets",
               "Milwaukee Bucks","Brooklyn Nets" ,"Detroit Pistons","Houston Rockets" ,"New Orleans Pelicans",
               "Indiana Pacers","Memphis Grizzlies","Denver Nuggets","Los Angeles Clippers","Atlanta Hawks",
               "New York Knicks" ,"Miami Heat","Orlando Magic","Dallas Mavericks","Phoenix Suns" ,
               "Sacramento Kings","Utah Jazz", "Minnesota Timberwolves","San Antonio Spurs","Cleveland Cavaliers",
               "Toronto Raptors" ,"Boston Celtics","Philadelphia 76ers" ,"Golden State Warriors","Oklahoma City Thunder")
)

get_ws_data = function(year){
  ws_url = paste0("https://www.basketball-reference.com/play-index/psl_finder.cgi?request=1&match=single&type=totals&per_minute_base=36&per_poss_base=100&season_start=1&season_end=-1&lg_id=NBA&age_min=0&age_max=99&is_playoffs=N&height_min=0&height_max=99&year_min=",
                  year,"&year_max=",year,"&birth_country_is=Y&as_comp=gt&as_val=0&pos_is_g=Y&pos_is_gf=Y&pos_is_f=Y&pos_is_fg=Y&pos_is_fc=Y&pos_is_c=Y&pos_is_cf=Y&order_by=ws")
  ws_stats = ws_url %>% read_html %>%
    html_table() %>%
    .[[1]] 
  ws_stats = ws_stats[,c(2,5,7)]
  colnames(ws_stats)= ws_stats[1,]  
  ws_stats = ws_stats[-c(1,which(ws_stats$Player=='Player')),]
  ws_stats$WS = as.numeric(ws_stats$WS )
  ws_stats_summary = ws_stats %>%
    group_by(Tm) %>%
    summarise(WS = max(WS))
  return(ws_stats_summary)
}
WinShare = get_ws_data(2018)
head(WinShare)


get_team_year_feature = function(year){
  
  #team_stats = get_team_stats() # get the team stats for all seasons
  team_stats_year = team_stats %>%
    filter(grepl(as.character(year-1),Season)) # filter the specific year 
  team_stats_year[,-c(2,3)] = as.numeric(as.matrix(team_stats_year[,-c(2,3)])) # turn characters into numeric
  
  
  ## select Win, Loss, Win-Loss per, Team points, Opponent points
  team_stats_year_feature = team_stats_year %>%
    select(Tm, W, L, `W/L%`, TeamPTS, OpponentPTS) %>%
    inner_join(west_ind, by = 'Tm')
  
  ## assign the team rank
  team_stats_year_rank = team_stats_year_feature %>%
    group_by(western) %>%
    mutate(ConfRank = order(order(W, TeamPTS, decreasing = T)))
  
  ## add the winshare feature from previous season
  Winshare  = get_ws_data(year-1)
  team_stats_year_rank = team_stats_year_rank %>% left_join(WinShare, by = 'Tm')
  team_stats_year_rank$WS[is.na(team_stats_year_rank$WS)] = 0
  
  return(team_stats_year_rank)
  
}

team_stats_18to19 = get_team_year_feature(2019)
head(team_stats_18to19)

get_game_year_feature = function(year){
  game_stats_year = game_stats %>% 
    filter(grepl(as.character(year-1),Season))
  ## game stats 
  ## the results of every game in the season 
  game_stats_feature = game_stats_year[,c(1,5,6,3,4)] 
  game_stats_feature[,c(3,5)] = as.numeric(as.matrix(game_stats_feature[,c(3,5)]))
  colnames(game_stats_feature) = c("Date","Home","HomePTS","Away","AwayPTS")
  game_stats_feature = game_stats_feature %>%
    mutate(HomeWin = ifelse(HomePTS > AwayPTS,1,0))
  return(game_stats_feature)
}
game_stats_18to19 = get_game_year_feature(2019)
head(game_stats_18to19)
```



## Exploratory Data Analysis
### Team-level 
```{r,warning=F,message=F}
WL_record_cal = function(game_schedule,WL_record){
  for(i in 1:nrow(game_schedule)){
    if (game_schedule$HomeWin[i]==1){
      WL_record$W[which(WL_record$TeamName==game_schedule$Home[i])] = 
        WL_record$W[which(WL_record$TeamName==game_schedule$Home[i])] + 1
      WL_record$L[which(WL_record$TeamName==game_schedule$Away[i])] = 
        WL_record$L[which(WL_record$TeamName==game_schedule$Away[i])] + 1
    }
    
    if (game_schedule$HomeWin[i]==0){
      WL_record$L[which(WL_record$TeamName==game_schedule$Home[i])] = 
        WL_record$L[which(WL_record$TeamName==game_schedule$Home[i])] + 1
      WL_record$W[which(WL_record$TeamName==game_schedule$Away[i])] = 
        WL_record$W[which(WL_record$TeamName==game_schedule$Away[i])] + 1
    }
  }
  return(WL_record)
}

team_def_off_plot = function(year){
  team_stats_year = get_team_year_feature(year)
  p = team_stats_year %>% 
    ggplot(aes(x = TeamPTS/20, y = OpponentPTS/20))+
    geom_point(aes(size = -ConfRank, col = factor(western)))+
    geom_text(aes(label=Tm),vjust = 0, nudge_y = 0.5)+
    geom_abline(slope = 1,intercept = 0,linetype="dashed", color = "red",size = 2)+
    facet_grid(~western)+
    labs(title = paste0('Season ',year-1,'-',substr(year,3,4)))+
    theme(plot.title = element_text(hjust = 0.5))+
    coord_fixed()+
    theme(legend.position = "bottom")
  print(p)
}
team_def_off_plot(2019)
```

```{r,warning=F,message=F}
team_rank_plot = function(year, Nfirst = 300){
  team_stats_year = get_team_year_feature(year)
  game_stats_year = get_game_year_feature(year)
  colnames(team_stats_year)[c(2,3)] = c("W.pre","L.pre")
  
  WL_record = cbind(west_ind,W = 0, L=0)
  WL_record  = WL_record_cal(game_stats_year,WL_record)
  colnames(WL_record)[c(4,5)] = c("W.all","L.all")
  WL_record = WL_record %>%
    group_by(western) %>%
    mutate(Season_Rank = order(order(W.all, decreasing = T))) %>%
    inner_join(team_stats_year[,c(1,2,3,8,9)],by=c("Tm","TeamName"))
  p = WL_record %>%
    ggplot(aes(x = ConfRank, y = Season_Rank))+
    geom_point(aes(col = factor(western)),size = 2)+
    geom_text(aes(label=Tm),vjust = 0, nudge_y = 0.5)+
    geom_abline(slope = 1,intercept = 0,linetype="dashed", color = "red",size = 2)+
    geom_hline(yintercept=8, linetype="dotted", color = "blue",size = 2)+
    facet_grid(~western)+
    labs(title = paste0('Season ',year-1,'-',substr(year,3,4)))+
    theme(plot.title = element_text(hjust = 0.5))+
    coord_fixed()+
    theme(legend.position = "bottom")
  print(p)
  
  p2 = WL_record %>%
    ggplot(aes(x = W.pre, y =W.all))+
    geom_point(aes(col = factor(western)),size = 2)+
    geom_smooth(method = "lm")+ 
    geom_text(aes(label=Tm),vjust = 0, nudge_y = 0.5)+
    facet_grid(~western)+
    labs(title = paste0('Season ',year-1,'-',substr(year,3,4)))+
    theme(plot.title = element_text(hjust = 0.5))+
    theme(legend.position = "bottom")
  print(p2)
}

team_rank_plot(2019)
```

### Specific effect
#### Home-effect
#### Western-Eastern effect
```{r}
home_away_plot = function(year){
  game_stats_year = get_game_year_feature(year)
  standing_url = paste0("https://www.basketball-reference.com/leagues/NBA_",
                        year,"_standings.html")
  
  standing_year = standing_url %>% 
    read_html %>%
    html_nodes(xpath = '//comment()') %>%    # select comment nodes
    html_text() %>%    # extract comment text
    paste(collapse = '') %>%    # collapse to a single string
    read_html() %>%    # reparse to HTML
    html_table() %>%
    .[[1]]  
  
  colnames(standing_year) = paste0(colnames(standing_year),standing_year[1,])
  standing_year = standing_year[-1,]
  
  standing_table = standing_year %>%
    select(Team,Overall,PlaceHome,PlaceRoad,ConferenceE,ConferenceW) %>%
    separate(PlaceHome,into = c("HomeW","HomeL"),sep = "-") %>%
    separate(PlaceRoad,into = c("AwayW","AwayL"),sep = "-") %>%
    separate(ConferenceE,into = c("EastW","EastL"),sep = "-") %>%
    separate(ConferenceW,into = c("WestW","WestL"),sep = "-") %>%
    separate(Overall,into = c("Wtot","Ltot"),sep = "-") %>%
    inner_join(west_ind,by=c("Team" = "TeamName"))
  
  standing_table_gather = standing_table %>%
    gather(stats,win,-c("Team","Wtot","Ltot","Tm","western"))
  standing_table_gather[,c(2,3,7)] = as.numeric(as.matrix(standing_table_gather[,c(2,3,7)]))
  p = standing_table_gather %>% 
    filter(stats %in% c("HomeW","AwayW")) %>%
    ggplot(aes(x = reorder(Tm,Wtot), y = win,fill= stats))+
    geom_bar(stat = "identity",position = "dodge")+
    facet_grid(~western,scales="free_x")+
    labs(title = paste0('Season ',year-1,'-',substr(year,3,4)))+
    theme(plot.title = element_text(hjust = 0.5))+
    theme(legend.position = "bottom")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
  print(p)
  
  team_factor_levels= standing_table_gather %>% 
    filter(western==1 & stats %in% c('EastW') | western==0 & stats %in% c('WestW')) %>%
    arrange(win) %>%
    select(Tm) %>% as.matrix() %>% as.character()
  
  p2 = standing_table_gather %>%
    mutate(Tm = factor(Tm,levels = team_factor_levels)) %>%
    filter(western==1 & stats %in% c('EastW','EastL') | western==0 & stats %in% c('WestW','WestL')) %>%
    ggplot(aes(x =Tm, y = win,fill= stats))+
    geom_bar(stat = "identity")+
    labs(title = paste0('Season ',year-1,'-',substr(year,3,4)))+
    theme(plot.title = element_text(hjust = 0.5))+
    theme(legend.position = "bottom")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
  print(p2)
  
  
}

home_away_plot(2019)
```


### Player-level
```{r}
team_ws_plot = function(year){
  team_stats_year = get_team_year_feature(year)
  game_stats_year = get_game_year_feature(year)
  colnames(team_stats_year)[c(2,3)] = c("W.pre","L.pre")
  
  WL_record = cbind(west_ind,W = 0, L=0)
  WL_record  = WL_record_cal(game_stats_year,WL_record)
  colnames(WL_record)[c(4,5)] = c("W.all","L.all")
  WL_record = WL_record %>%
    group_by(western) %>%
    mutate(Season_Rank = order(order(W.all, decreasing = T))) %>%
    inner_join(team_stats_year[,c(1,2,3,8,9)],by=c("Tm","TeamName"))
  
  
  WinShare = get_ws_data(year)
  WL_record_ws = WL_record %>%
    inner_join(WinShare, by='Tm')
  
  p = WL_record_ws %>%
    ggplot(aes(x = WS, y = Season_Rank))+
    geom_point(aes(col = factor(western)),size = 2)+
    geom_text(aes(label=Tm),vjust = 0, nudge_y = 0.5)+
    facet_grid(~western)+
    labs(title = paste0('Season ',year-1,'-',substr(year,3,4)))+
    theme(plot.title = element_text(hjust = 0.5))+
    theme(legend.position = "bottom")
  print(p)
}

team_ws_plot(2019)
```

```{r}
get_ws_raw_data = function(year){
  ws_url = paste0("https://www.basketball-reference.com/play-index/psl_finder.cgi?request=1&match=single&type=totals&per_minute_base=36&per_poss_base=100&season_start=1&season_end=-1&lg_id=NBA&age_min=0&age_max=99&is_playoffs=N&height_min=0&height_max=99&year_min=",
                  year,"&year_max=",year,"&birth_country_is=Y&as_comp=gt&as_val=0&pos_is_g=Y&pos_is_gf=Y&pos_is_f=Y&pos_is_fg=Y&pos_is_fc=Y&pos_is_c=Y&pos_is_cf=Y&order_by=ws")
  ws_stats = ws_url %>% read_html %>%
    html_table() %>%
    .[[1]] 
  ws_stats = ws_stats[,c(2,3,5,7)]
  colnames(ws_stats)= ws_stats[1,]  
  ws_stats = ws_stats[-c(1,which(ws_stats$Player=='Player')),]
  ws_stats$WS = as.numeric(ws_stats$WS )
  return(ws_stats)
}




player_ws_plot = function(){
  year_list = c(2015:2019)
  ws_stats  = NULL
  for(i in 1: length(year_list)){
      ws_tmp = get_ws_raw_data(year_list[i])
      ws_stats = rbind(ws_stats, ws_tmp)
  }
  ws_stats = ws_stats %>% arrange(-WS)
  player_rank_list = unique(ws_stats$Player)
  
  ws_stats_rank = ws_stats %>%
    group_by(Season) %>%
    mutate(WSRank = rank(-WS,ties.method = "first"))
  
  p = ws_stats_rank %>%
    filter(Player %in% player_rank_list[1:20]) %>%
    ggplot(aes(x = Player, y = WSRank,fill = Player))+
    geom_boxplot()+
    #labs(title = player_rank_list[player_rank])+
    theme(plot.title = element_text(hjust = 0.5))+
    theme(legend.position = "none")+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  print(p)
}
player_ws_plot()

```

## Feature Engineering
The performance of a team could be captured by both the team-level and player-level statistics. The team-level features that could be included in our current model were:
* number of W and L, W/L percentage, conference rank
* team total/average points, oppponent total/average points
* other secondary statistics

The player-level features that could be incorporated in the team performance metrics were

* number of league top (50-100) win share players in the team
* highest win share value in the team

## Data Analysis

### Statistical Models: Bradley-Terry Logistic Regression Model
```{r}
library(caret)

get_train_test_data = function(year, firstN){
  team_stats_year = get_team_year_feature(year)
  game_stats_year = get_game_year_feature(year)
  
  train_set = game_stats_year[-c(1:firstN),]
  team_per_diff = data.frame(team_stats_year[match(train_set$Home,team_stats_year$TeamName),c(2,5,6,7,9,10)]
                             - team_stats_year[match(train_set$Away,team_stats_year$TeamName),c(2,5,6,7,9,10)])
  train_set_teamper = data.frame(train_set,
                                 team_per_diff = I(as.matrix(team_per_diff)))
  
  return(train_set_teamper)
}



build_logit_model_year = function(train_year = 2018, test_year = 2019, firstN = 300){
  
  train_set = get_train_test_data(train_year,firstN)
  test_set = get_train_test_data(test_year,firstN)
  
  ## logit model
  logit_model = glm(HomeWin~ team_per_diff,family=binomial(link='logit'),
                    data = train_set)
  print(summary(logit_model)) # model summary
  
  ## training accuracy
  
  print(confusionMatrix(factor(ifelse(logit_model$fitted.values>0.5,1,0)),
                               factor(train_set$HomeWin)))
  
  ## testing accuracy
  pred_win_prob = predict(logit_model,newdata = test_set, type = "response")
  pred_game = ifelse(pred_win_prob>0.5,1,0)
  
  test_set_pred = cbind(test_set,pred_win_prob,pred_game)
  print(confusionMatrix(factor(test_set_pred$pred_game),factor(test_set_pred$HomeWin)))
  return(test_set_pred)
}

test_set_pred = build_logit_model_year(2017,2018)
head(test_set_pred)
```


### Machine Learning Models: Decision Tree, Random Forest, Neural Network

### Simulation Framework
```{r}
## Remaining 60 games
set.seed(2019)
B = 100
game_simulation = function(pred_result, B, year, Nfirst = 300){
  game_stats_year = get_game_year_feature(year)
  pretrain_set = game_stats_year[1:Nfirst,]
  train_set = game_stats_year[-c(1:Nfirst),]
  WL_record = cbind(west_ind,W = 0, L=0)
  
  WL_record_pre  = WL_record_cal(pretrain_set,WL_record)
  WL_record_post = WL_record_cal(train_set,WL_record)
  WL_record_all  = WL_record_cal(train_set,WL_record_pre)
  WL_record_west = WL_record_all %>%
    filter(western==1) %>%
    arrange(-W)
  
  
  WL_simu_rep = NULL
  for(k in 1:B){
    simu_game = rep(0,nrow(pred_result))
    
    for(i in 1:nrow(pred_result)){
      simu_game[i] = rbinom(1,1, pred_result$pred_win_prob[i])
    }
    simu_result = pred_result %>%
      select(Home,Away,HomeWin) %>%
      mutate(HomeWin = simu_game)
    
    # WL_record = cbind(west_ind,W = 0, L= 0)
    WL_simu = WL_record_cal(simu_result,WL_record_pre)
    WL_simu_west = WL_simu %>%
      filter(western==1) %>%
      mutate(TeamRank = rank(-W,ties.method = "random")) %>%
      mutate(playoff = ifelse(TeamRank<=8,1,0))
    WL_simu_rep = rbind(WL_simu_rep, WL_simu_west)
  }
  ans = list(simu = WL_simu_rep, result = WL_record_west)
  return(ans)
}
simu_result = game_simulation(test_set_pred, B, year = 2018)
train_set_simu = simu_result$simu
train_set_summary = train_set_simu %>%
  group_by(Tm) %>%
  summarise(W_avg = mean(W),
            L_avg = 82 - W_avg,
            W_med = median(W),
            L_med = 82 - W_med,
            playoff_prob = sum(playoff)/B) %>%
  arrange(-W_avg)
print(train_set_summary)

WL_record_west = simu_result$result
print(WL_record_west)
```


### Extension: An Ensemble Learning Method for Integrative Prediction Using Mutiple Seasons Data

## Summary 
